Schema Evolution in Interactive Programming Systems

### 2024-08-28 paper.tex
simplify wording
move citation before period

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

>>>>

In this paper we identify a common problem, \emph{Schema Change}, that interferes with feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations ~\cite{erhard06}. We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.
